Network based multifactorial modelling microRNA-target interaction.

### 2019-04-11 article.Rmd
article rev

---
title: "Network based multifactorial modelling microRNA-target interaction."
author: 
- Selcen Ari^[Yildiz Technical University Department of Bioengineering]
- Alper Yilmaz^[Yildiz Technical University Department of Bioengineering]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_document: default
bibliography: ref.bib
biblio-style: NAR-natbib.sty
---

@@

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d).

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions have observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation. On the other hand, It was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors.This also shows that the perturbation iniated element in the system is important parameter. So, we also have developed an approach that is found perturbation efficiencies of elements in dataset. **(see details in Supplementary File(fig1_2app.Rmd)**. When we applied the method on the minimal dataset, while Gene4 is the most efficient element in terms of number of perturbed elements, it have observed that Mir2 is most potent element in the mean of expression change.

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Firstly, we have observed the activities in the iterations and determined behaviours of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviours and complexity to gain steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibit competing behaviours in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, somes have changed slightly. So, we have compared two datasets: the tumor gene expression and simulation results. When we analysed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.

The network based approaches have been developed in previous studies. Figliuzzi et al. have tried to explain the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs. They have pointed out that the larger miRNA number can cause to evident crosstalk between ceRNAs. It has been thought that miRNAs and their targets interactions depend on rates of transcription, degradation, binding and unbinding in the network based kinetic model which has been developed by Nitzan et al. Nitzan and collaborators have used the high throughput experimental dataset [@helwak_mapping_2013] for miRNA:target pair that occur high free energy and microarray dataset of a miR-92a depletion for expression values. They have been demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions change to depend on distance between ceRNAs, and topological features of network [@nitzan_interactions_2014].

>>>>

---
title: "Network based multifactorial modelling microRNA-target interaction."
author: 
- Selcen Ari^[Yildiz Technical University Department of Bioengineering]
- Alper Yilmaz^[Yildiz Technical University Department of Bioengineering]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: 
  html_document: default
bibliography: ref.bib
---

@@

Competing endogeneous RNA (ceRNA) regulations and crosstalk in human are remarkable issue in mean of regulations via miRNAs. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected regulations because of indirect and complex interaction. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values and interaction factors optionally. We applied approach in sample dataset and defined dynamics of it. We have observed that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be shown. Importance of miRNA:target ratios was also observed consistently with previous reports. We have obtained that the competing elements which have the same or close expression values may not affected equal from the perturbation because of repression functionality of interactions. To apply model to the real sample that obtained combining  breeast cancer gene and miRNA expression dataset via high throughput experimental miRNA:target dataset, we selected a gene, *ABCC1*, which over expressed in tumor tissue and utilized it as perturbating element. We have observed that one change can be sufficient to perturb regulations, and may not predicted all of regulations correctly, with this change. Therefore, this model which helps to understand crosstalking between elements could be tested and improved via results of well-constructed experimental dataset and takes to hope for future research.

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. But these factors can be handled optionally and differently.

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions have observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation. On the other hand, It was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors.This also shows that the perturbation-initiation element in the system is important parameter. So, we also have developed an approach that is found perturbation efficiencies of elements in dataset. **(see details in Supplementary File(fig1_2app.Rmd)**. When we applied the method on the minimal dataset, while Gene4 is the most efficient element in terms of number of perturbed elements, it have observed that Mir2 is most potent element in the mean of expression change. In our approach, the elements behave differently based on type (competing or miRNA). When a competing element was changed, whole elements out of miRNAs are re-regulated. But, if a miRNA was perturbed, whole competing elements are re-regulated and miRNA preserves expression at after perturbation.

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Firstly, we have observed the activities in the iterations and determined behaviours of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviours and complexity to gain steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibit competing behaviours in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, somes have changed slightly. So, we have compared two datasets; the tumor gene expression and simulation results. When we analysed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene and miRNA expression values at initial conditions. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.

The network based approaches have been developed in previous studies. Figliuzzi et al. have tried to explain the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs. They have pointed out that the larger miRNA number can cause to evident crosstalk between ceRNAs. It has been thought that miRNAs and their targets interactions depend on rates of transcription, degradation, binding and unbinding in the network based kinetic model which has been developed by Nitzan et al. Nitzan and collaborators have used the high throughput experimental dataset [@helwak_mapping_2013] for miRNA:target pair that occur high free energy and microarray dataset of a miR-92a depletion for expression values. They have been demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions change to depend on distance between ceRNAs, and topological features of network [@nitzan_interactions_2014]. List et al. have developed an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets [@markus_list_sponge_2017]. Their approach could be useful to analysis interacted genes through miRNAs. When it has been throught that a miRNA can exhibit strong functionality to a target but may not against an other, common miRNA based approach can be inconvenient to understand regulations of ceRNA interactions.
